---
title: "Practical Machine Learn-Project Write Up"
author: "Andria Hall"
date: "Sunday, October 26, 2015"
output: html_document
---

### Background

This project write-up is to build a machine learning model based on a group of enthusiasts who take measurements about themselves regularly to improve their health. This data comes from accelerometers on the belt, forearm, arm and dumbell of **six** participants who were asked to perform barbell lifts correctly and incorrectly in five different ways.

### Reproduceability

We will load the appropriate packages and set the seed for reproduceable results.

```{r load_packages}
library(randomForest)
library(parallel)
library(doParallel)
library(rattle)
library(rpart.plot)
```

We will now import the testing data and the training data and test for identicality.

```{r}
library(data.table)

# Load the training data set
setInternet2(TRUE)

# Load the training data set
url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
DTrain <- fread(url)
trainColnames <- colnames(DTrain)

# Load the testing data set
url <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
DTest <- fread(url)
testColnames <- colnames(DTest)

# Verify that the column names (excluding classe and problem_id) are identical in the training and testing set.
all.equal(trainColnames[1:length(trainColnames)-1], testColnames[1:length(trainColnames)-1])
```

We will now remove all `NA` values and columns: `X`, `user_name`, `raw_timestamp_part_1`, `raw_timestamp_part_2`, `cvtd_timestamp`, `new_window` `num_window` as they are not predictors.

```{r}
# Discard columns with missing values
DiscardNAs <- sapply(DTest, function (x) any(is.na(x) | x == ""))
isPredictor <- !DiscardNAs & grepl("belt|[^(fore)]arm|dumbbell|forearm", names(DiscardNAs))
predData <- names(DiscardNAs)[isPredictor]
predData
```


We will now subset the predData to include only the predictors and the outcome variable, `classe`.

```{r}
# Subset predData dataset
predDataClasse <- c("classe", predData)
DTrain <- DTrain[, predDataClasse, with=FALSE]
dim(DTrain)
names(DTrain)
```

and make `classe` a factor.

```{r}
DTrain <- DTrain[, classe := factor(DTrain[, classe])]
DTrain[, .N, classe]
```

We will next Split the DTrain into 60% training and 40% testing dataset.


```{r}
library(caret)
seed <- as.numeric(as.Date("2014-10-26"))
set.seed(seed)
trainIndex <- createDataPartition(DTrain$classe, p=0.6)
trainData <- DTrain[trainIndex[[1]]]
testData <- DTrain[-trainIndex[[1]]]
```

Preprocess the prediction variables by centering and scaling.

```{r}
X <- trainData[, predData, with=FALSE]
preProc <- preProcess(X)
preProc

XCS <- predict(preProc, X)
DTrainCS <- data.table(data.frame(classe = trainData[, classe], XCS))
```

Apply the centering and scaling to the probing dataset.

```{r}
X <- testData[, predData, with=FALSE]
XCS <- predict(preProc, X)
DProbeCS <- data.table(data.frame(classe = testData[, classe], XCS))
```

We will check for covariates that have near zero variance

```{r}
nzv <- nearZeroVar(DTrain, saveMetrics=TRUE)
if (any(nzv$nzv)) nzv else message("No variables with near zero variance")
```

Training the dataset using random forest algorithm


```{r}
# Configure the train control for cross-validation

cl <- makeCluster(detectCores() -1)
registerDoParallel(cl)

tctrl <- trainControl(classProbs=TRUE, savePredictions=TRUE, allowParallel = TRUE)

# Fit the model using Random Forests algorithm
Fitmod <- randomForest(classe ~.,  data = DTrainCS,
                method="rf")
stopCluster(cl)
```

```{r}
print(Fitmod)
```

### Evaluate the model on the training dataset

```{r}
pred <- predict(Fitmod, DTrainCS)
confusionMatrix(pred, DTrainCS[, classe])
```

### Evaluate the model on the testing dataset

```{r}
pred <- predict(Fitmod, DProbeCS)
confusionMatrix(pred, DProbeCS[, classe])
```

### Display the final model

```{r}
varImp(Fitmod)
```

Examine groups of predictable variable

```{r}
set.seed(1)
modFit <- rpart(classe ~ ., data=DTrainCS, method="class")

fancyRpartPlot(modFit)
```


Save training model object for later.

```{r}
save(Fitmod, file="Fitmod.RData")
```


### Predict on the test data

```{r}
# Load Fitmod data
load(file="Fitmod.RData", verbose=TRUE)
```

Get predictions and evaluate

```{r}
DTestCS <- predict(preProc, DTest[, predData, with=FALSE])
pred <- predict(Fitmod, DTestCS)
DTest <- cbind(pred, DTestCS)
subset(DTest, select=names(DTest)[grep("belt|[^(fore)]arm|dumbell|forearm", names(DTest), invert=TRUE)])
```

### Write submission files to `pred` in Practical Machine Learning folder 

```{r}
pml_write_files = function(x){
  n = length(x)
    for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
pml_write_files(pred)
```